---
title: "Figure 2"
author: "Alisa Greenberg"
date: "April 2024"
output:
  html_document:
    css: custom_notebook_formatting.css
    toc: true
    toc_depth: 3
    df_print: paged
    code_folding: hide
    number_sections: yes
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.width=6, fig.height=4) 
```

```{r warning=FALSE, message=FALSE, include=FALSE}
library("dplyr")
library("ggplot2")
library("readr")
library("tidyr")
library("pROC") 

theme_set(theme_bw())
files.sources = file.path("R",list.files(path = "R"))
sapply(files.sources, source)

response_colors <- c( "No response" = "#CB5F8A", "Response" = "#0080D2")
feat_table_dir <- "feature_tables"
```

# Data import and preparation
```{r}
# Data import
extract_metadata_from_sample_name <- function(data, sample_var = "sample"){
  data <- data %>% 
          mutate(subject = stringr::str_extract(!!sym(sample_var),"[0-9]+"),
                 subject = paste0("subject_",subject),
                 timepoint = stringr::str_extract(!!sym(sample_var), "PRE|POST"),
                 timepoint = factor(tolower(timepoint),levels = c("pre", "post")))
  return(data)
}
```

```{r}
## Metadata
metadata <- read.csv("metadata/metadata.csv") %>% 
  mutate(response = ifelse(response == "response", "Response", "No response"))
```

```{r}
## Select key features for downstreatm analysis
features <- c("subject", "response", "tdee_adaptation", "rmr_adaptation")
metadata_subject_level <- metadata %>% select(features)
```

```{r}
sample_mapping <- rbind(metadata_subject_level %>% mutate(timepoint = "PRE"),
                        metadata_subject_level %>% mutate(timepoint = "POST"))
sample_mapping <- sample_mapping %>%
                  mutate(sample = paste(subject, timepoint, sep = "_"),
                         subject_number = gsub("subject_", "", subject),
                         sample_id = paste("ARE",subject_number,timepoint,
                                           "CP04693_L001", sep = "_"),
                         response = factor(response),
                         timepoint = factor(timepoint, levels = c("PRE","POST"))) %>%
                  filter(sample != "subject_15_POST") # Missing sample
```

## Figure 2
```{r}
## Relative abundance 
kraken_sp <- read_tsv2("feature_tables/kraken_species_level_taxonomy.tsv")
kraken_sp <- reorganize_kraken_table(kraken_sp)
ra_kraken_sp <- convert_to_relab(kraken_sp)

ra_kraken_dfs <- list("species"=ra_kraken_sp)
```

SUbset the data for species with sufficient amount of samples
```{r}
ra_kraken_sp_baseline <- ra_kraken_sp %>% filter(sample_id %in% grep("PRE",ra_kraken_sp$sample_id, value = T))
min_n_samples <- (nrow(metadata) * 0.8) %>% round()
sp_with_sufficient_samples <- colSums(ra_kraken_sp_baseline %>% select(-sample_id) != 0)
sp_with_sufficient_samples <- sp_with_sufficient_samples[sp_with_sufficient_samples > min_n_samples] %>% names()

sp_mean_abundance <- ra_kraken_sp_baseline %>% 
                    select(sp_with_sufficient_samples) %>% 
                    apply(., 2, function(x) mean(x, na.rm = TRUE)) %>% 
                    data.frame() %>% rename(mean_abundance = ".") %>%
                    tibble::rownames_to_column("species") 

sp_with_sufficient_samples <- sp_mean_abundance %>% filter(mean_abundance > 0.0005)
```

```{r}
gtdb_species <- c(grep("Faecalibacterium.prausnitzii",sp_with_sufficient_samples$species, value = TRUE))
ncbi_targets <- c("Faecalibacterium.sp003449675","Faecalibacterium.sp900539885","Faecalibacterium.sp900540455",
                  "Faecalibacterium.sp905215595")
target_species <- intersect(sp_with_sufficient_samples$species, c(gtdb_species, ncbi_targets))

target_species_ra <- ra_kraken_sp %>% select(sample_id, target_species) %>% 
                     extract_metadata_from_sample_name("sample_id")

target_species_df <- metadata_subject_level %>%
                     left_join(metadata %>% select(subject, tef_4hours_pre, rer_pre), by = c("subject")) %>%
                     left_join(target_species_ra , by = c("subject"))
```


```{r}
species_ra <- ra_kraken_sp  %>% 
              select(sample_id, sp_with_sufficient_samples$species) %>% 
              extract_metadata_from_sample_name("sample_id") %>% filter(timepoint == "pre")

data <- metadata_subject_level %>% left_join(species_ra , by = c("subject")) %>% 
              mutate(response = factor(response, levels = c("Response", "No response")))
```

```{r message=FALSE, warning=FALSE}
feature_vec <- c()
auc_vec <- c()
roc_df <- data.frame()
random_seed <- 100

features <- sp_with_sufficient_samples$species
for (feature_name in features){
  feature_vec <- c(feature_vec, feature_name)
  cv_roc <- calculate_cv_roc(data, 
                             features = c(feature_name),
                             target = "response",
                             n_ml_shuffle = 0,
                             n_ml_folds = 3,
                             rand_seed = random_seed)
  
  roc <- cv_roc[["roc"]] %>% mutate(feature = feature_name)
  roc_df <- bind_rows(roc_df, roc)
  
  auc <- cv_roc$auc[["auc"]]
  auc_vec <- c(auc_vec, auc)
}
auc_df <- data.frame(feature = feature_vec, auc = auc_vec) %>%
          mutate(target_flag = ifelse(feature %in% target_species, "Target", "Other"),
                  is_highly_predictive = auc > 0.8)
roc_df <- roc_df %>% mutate(target_flag = ifelse(feature %in% target_species, "Target", "Other"),
                            is_highly_predictive = auc > 0.8)
# write.csv(auc_df, "plots/auc_response_by_species_abundanc.csv")
```

### Statistics - test enrichment of F.Prausnitzii subspecies in the top of the species ranking by ROCAUC

Calculate median, mean and starndard deviation of species groups ROCAUC
```{r}
auc_df %>% group_by(target_flag) %>% summarise(median = median(auc), mean= mean(auc), SD = sd(auc))
```

Check the enrichment of target species at the top of the AUC ranks
```{r}
auc_df <- auc_df %>% arrange(desc(auc))
target_ranks <- match(target_species, auc_df$feature) 
enrichement_pval <- gsEasy::gset(S = target_ranks, N = nrow(auc_df))

enrichement_pval_Other <- replicate(n=10, expr=gsEasy::gset(S=sample.int(n=nrow(auc_df), size=length(target_species)), N=nrow(auc_df)))
target_auc <- auc_df %>% filter(target_flag == "Target") %>% pull(auc)
Other_auc <- auc_df %>% filter(target_flag != "Target") %>% pull(auc)
wilcox.test(target_auc,  Other_auc)
```

### Figure 2A - ROC curves for response prediction based on single-species pre-training relative abundance
```{r}
ggdata <- auc_df %>% filter(is_highly_predictive) %>% 
          mutate(name_for_plotting = gsub("CAG.303", "CAG_303",feature),
                 name_for_plotting = gsub("\\.", "\n",name_for_plotting))

species_palette <- c(
  "black",
"#00827D",
"#0080D2",
"#5D74D0",
"#AA64B3",
"#CB5F8A",
"#ED727F",
"#FF8D71",
"#FFAE65")
names(species_palette) <- ggdata$name_for_plotting
```

```{r}
list_curves <- list()
for (g in ggdata$feature){
  roc_df <- calculate_cv_roc(data, c(g), "response", n_ml_shuffle = 0, n_ml_folds = 3,rand_seed = random_seed, smooth_roc =F)
  list_curves[[g]] <- roc_df$rocs[["0"]]
}
names(list_curves) <- ggdata$name_for_plotting
ggp <- ggroc(list_curves,size = 1.5, alpha = 0.7) +  
  annotate("segment",x = 1, xend = 0, y = 0, yend = 1, color="black", linetype="dashed") +
  scale_color_manual(values = species_palette) +
  labs(color = "Species") + #theme(text = element_text(family = "Outfit",size = 12)) +
  custom_theme()
plot(ggp)
#ggsave(plot = ggp, filename = "plots/top_species_roc_curves.png", height = 5, width = 5)
```

### Figure 2B - ROCAUC values for response prediction based on single-species pre-training relative abundance
```{r}
ggp <-  ggplot(ggdata, aes(reorder(name_for_plotting,auc), auc, fill = name_for_plotting)) + 
        geom_col(alpha = 0.8) + coord_flip() + 
        scale_fill_manual(values = species_palette) +
        labs(y = "AUROC", x = "") + custom_theme() + 
        ylim(0,1) + geom_hline(yintercept = 0.5, linetype = "dashed")
plot(ggp)
#ggsave(plot = ggp, filename = "plots/top_species_auc_bar.png", height = 3.5, width = 4)
```

### Figure 2S 

Test if the relative abundances of highly predictive species are significantly different between response groups pre-training
```{r}
target_species <- auc_df %>% filter(is_highly_predictive) %>% pull(feature)
p_value <- c()

for (g in target_species){
  wilcox <- data %>% wilcox.test(as.formula(paste0(g,"~response")), data = ., paired = FALSE)
  p_value <- c(p_value, wilcox$p.value)
}
res <- data.frame(target_species, p_value) %>% mutate(p_value_adj = p.adjust(p_value, "fdr"))
res
```

Plot relative abundance values of highly predictive species between response groups pre-training
```{r}
data <- data %>% mutate(response = as.character(response))
ggp_list <- list()
p_value <- c()
name <- c()
top_species <- auc_df %>% filter(is_highly_predictive) %>% pull(feature)
for (g in top_species){
  name <- c(name, g)
  wilcox <- data %>% wilcox.test(as.formula(paste0(g,"~response")), data = ., paired = FALSE)
  p_value <- c(p_value, wilcox$p.value)

subtitle <- g %>% gsub("\\.sp", "\nsp", .) %>% gsub("\\.prausnitzii", "\nprausnitzii",.)
ggp_list[[g]] <- geom_boxplot_general(data, "response",  g, "response", response_colors) +
      labs(x = "", y = "Relative abundance", subtitle = subtitle) + scale_y_log10() + 
      custom_theme(is_figure_small = T)#theme(legend.position = "none")
}
```


```{r,  results='asis', fig.width=10, fig.height= 10}
margin = theme(plot.margin = unit(c(0.15,0.25,0.2,0.2), "cm"))
ggp<- gridExtra::grid.arrange(grobs = lapply(ggp_list, "+", margin),ncol = 3)
ggsave(plot = ggp, filename = "plots/top_predictive_species_RA.png", height = 9, width =9)

res <- auc_df %>% filter(is_highly_predictive) %>% 
       mutate(p_value = p_value,
              p_value_adj = round(p.adjust(p_value, "fdr"),4))
res %>% filter(p_value_adj < 0.05)
```