---
title: "Figure 1"
author: "Alisa Greenberg"
date: "April 2025"
output:
  html_document:
    css: custom_notebook_formatting.css
    toc: true
    toc_depth: 3
    df_print: paged
    code_folding: hide
    number_sections: yes
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.width=6, fig.height=4) 
```

```{r warning=FALSE, message=FALSE}
library("dplyr")
library("ggplot2")
files.sources = file.path("R",list.files(path = "R"))
sapply(files.sources, source)

theme_set(theme_bw(base_size = 12))
```

# plotting setup
```{r}
response_colors <- c("No response" = "#CB5F8A", "Response" = "#0080D2",
                     "pre" = "#b26636", "post" = "#00827D",
                     "Pre\ntraining" = "#b26636", "Post\ntraining" = "#00827D"
                     )
mass_type_colors <- c("Fat-free mass" = "#cb5f8a", "Fat mass" = "#FFAE65")
theme_set(theme_bw())
```

```{r}
label_dict <- c("pre" = "Pre training",
                "post" = "Post training",
                "response" = "Response", 
                "no_response" = "No response")
```


# Data loading
```{r}
metadata <- read.csv("metadata/metadata.csv") %>% 
            mutate(response = ifelse(response == "response", "Response", "No response"))
```

# Analysis 
## Figure 1 and S1
### Figure 1B - definition of response by change in FM and FFM
```{r}
ggdata <- metadata %>% 
          select(subject,response, gender, fm_kg_delta, ffm_delta) %>% 
          tidyr::pivot_longer(cols = c("fm_kg_delta", "ffm_delta")) %>% 
          mutate(name = case_when(name == "fm_kg_delta" ~ "Fat mass",
                                  name == "ffm_delta" ~ "Fat-free mass"),
                 subject =stringr::str_to_title(gsub("_"," ",subject)))
```

Subject weight changes (sorted by net. weight change)
```{r, }
ggp <- ggplot(ggdata, aes( x=reorder(subject, value,  function(x){ sum(x) }), y=value, fill=name)) + 
       geom_bar(position="stack", stat="identity", alpha = 0.7) + 
       scale_fill_manual(values = mass_type_colors) + 
       geom_hline(yintercept = 0, linetype = "dashed") + 
       labs(y = "Change from baseline [Kg]", x = "", 
            title = "Delta weight between post and pre training [Kg]") + 
       coord_flip()
plot(ggp + labs(subtitle = paste(strrep(" ", 10),"Mass loss",strrep(" ", 40) ,"Mass gain")))
```

Subject weight changes (sorted by response)
```{r}
ggp_facet <- ggp+ facet_wrap(~response, scales = "free_y") + custom_theme() +ggtitle("")
legend <- ggp %>% cowplot::get_legend() %>% ggpubr::as_ggplot()
ggp_facet
ggsave(ggp_facet, filename = "plots/response_barplot.png", width =7, height = 3.5)
ggsave(legend, filename = "plots/response_barplot_legend.png", width =2, height = 2)

```

### Figure 1C - weight pre vs post scatterplot
```{r}
ggp <- ggplot(metadata, aes(weight_pre, weight_post, color = response)) + 
       geom_abline(slope =1, intercept = 0, linetype = "dashed") +
       geom_point(size = 3, alpha = 0.5) + 
       scale_color_manual(values = response_colors) +
       labs(x= "Total weight pre-training [Kg]",
            y = "Total weight post-training [kg]",color = "Response\nto training") +
  theme(legend.key = element_blank(),
        legend.background=element_blank(),
        legend.position = c(0.8,0.15),
        legend.title.align=0.5) + custom_theme(is_figure_small = T, remove_legend = F)
plot(ggp)
ggsave(ggp, filename = "plots/weight_pre_post_scatter.png", width =3.5, height = 3.5)
```

```{r}
# Data preparation
for (feature in c("weight", "fm_kg", "ffm")){
ggdata <- metadata %>% 
          select(subject,response, paste(feature, c("pre", "post"), sep = "_")) %>% 
          tidyr::pivot_longer(cols = starts_with(feature),
                              names_to = "timepoint", values_to = feature) %>% na.omit()

p_r <- ggdata %>% filter(response == "Response") %>% 
            wilcox.test(as.formula(paste0(feature,"~timepoint")), data = ., paired = TRUE)
p_nr <- ggdata %>% filter(response == "No response") %>% 
             wilcox.test(as.formula(paste0(feature,"~timepoint")), data = ., paired = TRUE)

print(paste(feature,": Wilcoxon test - Response (pre vs post):", round(p_r$p.value,4)))
print(paste(feature,": Wilcoxon test - Non response (pre vs post):",  round(p_nr$p.value,4)))
}
```


### Figure 1D - VO2max differences between responders and non-responders, pre and post training
```{r}
# Data preparation
feature = "vo2max"
ggdata <- metadata %>% 
          select(subject,response, paste(feature, c("pre", "post"), sep = "_")) %>% 
          tidyr::pivot_longer(cols = starts_with(feature),
                              names_to = "timepoint", values_to = feature) %>% na.omit()

# For paired analysis, keep only subjects with 2 measurements (subject 15 is missing the post training measurement)
subject_missing_data <- ggdata %>% count(subject) %>% filter(n == 1) %>% pull(subject)

ggdata <- ggdata %>% filter(!subject %in% subject_missing_data) %>% 
          mutate(timepoint = gsub(paste0(feature,"_"), "", timepoint), 
                 timepoint = factor(timepoint, levels =  c("pre", "post")))
```

VO2max pre vs post (regardless of response group) - non significant
```{r}
wilcox <- ggdata %>% wilcox.test(as.formula(paste0(feature,"~timepoint")), data = ., paired = TRUE)
print(wilcox$p.value)

ggp <- geom_boxplot_general(ggdata, "timepoint",  feature, group_var = "subject") + 
  scale_x_discrete(labels = label_dict)+
  scale_color_manual(values = response_colors) + 
  scale_fill_manual(values = response_colors) + 
  labs(x = "Time point", y = "VO2 max") + 
  custom_theme(is_figure_small = T)
plot(ggp)
```

VO2max paired pre vs post by response group - significant increase only in responders
```{r}
# wilcoxon test by group
p_r <- wilcox.test(as.formula(paste0(feature,"~timepoint")),
                          data = ggdata %>% filter(response == "Response"), paired = TRUE)
p_nr <- wilcox.test(as.formula(paste0(feature,"~timepoint")),
                          data = ggdata %>% filter(response != "Response"), paired = TRUE)
print(paste("Wilcoxon test - Response (pre vs post):", p_r$p.value))
print(paste("Wilcoxon test - Non response (pre vs post):", p_nr$p.value))

# plot faceted by response
ggp_response <- ggp + facet_wrap(~response)
plot(ggp_response)
ggsave(ggp_response, filename = glue::glue("plots/{feature}_response_boxplot.png"), width =7, height = 3.5)

```

### Suplementory figure S1 - Pre-training values of possible confounders between response groups




Test differences in pre-training continues features across groups (FDR corrected wilcox rank-sum test)
```{r}
features <- c( "age","bmi_pre",  "ffm_pre", "fm_kg_pre","extime_perwk", "caloric_intake_pre", "caloric_intake_post")
feature_names <- c("Age [years]", "BMI pre training",
                   "FFM pre training [Kg]", "FM pre training [Kg]",
                   "Mean exercise time\n[min/week]", "Mean Caloric intake\npre-training [Kcal/day]",
                   "Mean Caloric intake\npost-training [Kcal/day]")
names(feature_names) <- features
```

```{r fig.height= 8, fig.width=9}
ggp_list <- list()
pvals <- c()

for(feature in features){
ggp_list[[feature]] <- geom_boxplot_general(metadata, "response",  feature, "response", response_colors) + 
      theme(legend.position = "none") + custom_theme(is_figure_small = T) + 
      labs(x = "", y= feature_names[[feature]]) #y= toupper(gsub("_pre","", feature)))

  wilcox <- metadata %>% wilcox.test(as.formula(paste0(feature,"~response")), data = ., paired = FALSE)
  pvals <- c(pvals, wilcox$p.value)
}
```
Caloric intake pre vs post
```{r, fig.height=3, fig.width=4.5}
ggdata <- metadata %>% select(subject, response, kcal_pre, kcal_post) %>% 
  tidyr::pivot_longer(cols = c("kcal_pre", "kcal_post"), names_to = "timepoint", values_to = "kcal") %>% 
  mutate(timepoint = factor(ifelse(timepoint == "kcal_pre", "pre-training", "post-training"), levels = c("pre-training", "post-training")))

for (r_status in c("Response", "No response")){
  data_subset <- metadata %>% filter(response == r_status)
  wilcox <- wilcox.test(data_subset$kcal_pre, data_subset$kcal_post, paired = TRUE)
  feature_names[paste("caloric intake pre-post", r_status)] <- paste("caloric intake pre-post", r_status)
  pvals <- c(pvals, wilcox$p.value)
}  
ggp <- geom_boxplot_general(ggdata, "timepoint",  "kcal", "response", response_colors) + 
  labs(x = "", y = "Mean Caloric intake [Kcal/day]") + 
  facet_wrap(~response) + theme(legend.position = "none") + custom_theme(is_figure_small = T)
ggp
#ggsave(ggp, filename = "plots/confounding_response_features_boxplot_addition.png", width =6, height = 3)
```
Test differences in categorical features (gender) by a Fisher test
```{r, fig.height=3, fig.width=3}
pval <- (metadata %>% select(gender, response) %>% table() %>% fisher.test())$p.value
pvals <- c(pvals, pval)
feature_names <- c(feature_names, "Gender")
ggp <- ggplot(metadata %>% count(gender, response),
       aes(gender,response, fill = n, label = n )) + 
       scale_fill_gradient(low = "#FFAE65", high = "#cb5f8a") + 
       geom_tile(alpha = 0.7) + geom_label(fill = "white") + custom_theme() + 
       labs(x = "Subject gender", y = "Response to training",
            fill = "# subjects")
ggp
#ggsave(ggp, filename = "plots/response_gender_fisher.png", width =2.5, height = 2.5)
```

```{r fig.height= 8, fig.width=9}
pvals_adj <- p.adjust(pvals, method = "fdr")
res <- data.frame(feature_names, pvals = round(pvals,2), fdr = round(pvals_adj,2))
res
ggp <- do.call(gridExtra::grid.arrange,  c(ggp_list, ncol = 3, top = ""))
#ggsave(ggp, filename = "plots/confounding_response_features_boxplot.png", width =9, height = 8)
```

