---
title: "Figures 3 and 4 - alpha and beta diversity measures"
author: "Alisa Greenberg"
date: "April 2025"
output:
  html_document:
    css: custom_notebook_formatting.css
    toc: true
    toc_depth: 3
    df_print: paged
    code_folding: hide
    number_sections: yes
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.width=6, fig.height=4) 
```

```{r warning=FALSE, message=FALSE, include=FALSE}
library("dplyr")
library("ggplot2")
library("readr")
library("tidyr")

theme_set(theme_bw(base_size = 12))
files.sources = file.path("R",list.files(path = "R"))
sapply(files.sources, source)

response_colors <- c( "No response" = "#CB5F8A", "Response" = "#0080D2")
feat_table_dir <- "feature_tables"
```

# Data import and preparation

```{r}
## Metadata
metadata <- read.csv("metadata/metadata.csv") %>% 
  mutate(response = ifelse(response == "response", "Response", "No response"))
```

```{r}
## Select key features for downstreatm analysis
features <- c("subject", "response", "tdee_adaptation", "rmr_adaptation", "vo2max_delta")
metadata_subject_level <- metadata %>% select(features)
```

```{r}
sample_mapping <- rbind(metadata_subject_level %>% mutate(timepoint = "PRE"),
                        metadata_subject_level %>% mutate(timepoint = "POST"))

vo2max_mes <- metadata %>% select(subject, vo2max_pre, vo2max_post) %>% 
              tidyr::pivot_longer(cols = -subject, values_to = "vo2max", names_to ="timepoint") %>% 
              mutate(timepoint = toupper(gsub("vo2max_","",timepoint)))

sample_mapping <- sample_mapping %>% left_join(vo2max_mes, by = c("subject", "timepoint")) %>% 
                  mutate(sample = paste(subject, timepoint, sep = "_"),
                         subject_number = gsub("subject_", "", subject),
                         sample_id = paste("ARE",subject_number,timepoint,
                                           "CP04693_L001", sep = "_"),
                         response = factor(response),
                         timepoint = factor(timepoint, levels = c("PRE","POST"))) %>%
                  filter(sample != "subject_15_POST") # Missing sample
```


```{r}
## Alpha diversity 
shannon <- read_delim("feature_tables/shannon.tsv", 
                      delim = "\t", escape_double = FALSE, 
                      col_names = c('sample_id', 'shannon'), 
                      skip = 1, trim_ws = TRUE, show_col_types = FALSE) %>% 
           extract_metadata_from_sample_name("sample_id")

## Beta diversity
bc <- read_tsv2("feature_tables/bc.tsv") %>% rename(sample_id = 1)
wuf <- read_tsv2("feature_tables/w_unifrac.tsv") %>% rename(sample_id = 1)

# Organize in lists for convenience
alpha_dfs <- list("shannon"=shannon)
distances_dfs <- list("Bray_Curtis"=bc, "Weighted_UniFrac"=wuf)
```

```{r}
## Setup feature names for plotting
feature_dict <- list("response" = "Response to training",
                     "tdee_adaptation" = "TDEE adaptation [Kcal]",
                     "rmr_adaptation" = "RMR adaptation [Kcal]",
                     "vo2max_delta" = "VO2 max delta")

metric_dict_alpha <- list("shannon" = "Shannon",
                          "shannon_delta" = "\u0394 Shannon")

metric_dict_beta <- list("Bray_Curtis" = "Bray Curtis",
                          "Weighted_UniFrac" = "Weighted UniFrac")
```

# Analysis 
## Figures 3A and S4 - beta diversity distances between response gorups and time points (PCoA)
Test if response groups create distinct clusters (rejecting H0 of the clusters having equivalent centroids and dispersion)

```{r}
pvals <- data.frame()
res_list <- list()
for (metric_name in names(distances_dfs)){
    ## Data setup
    ggdata <- sample_mapping
    samples <- ggdata$sample_id
    dist_matrix <- distances_dfs[[metric_name]] %>% 
                   filter(sample_id %in% samples) %>%
                   select(sample_id, samples)
    
    ## Calculation
    res <- get_permanova(disct_metadata = ggdata,  
                      dist_matrix = dist_matrix,
                      dist_metric_name = metric_name,
                      sample_id_column = 'sample_id', 
                      metadata_cols_to_analyze = c("response"),
                      fdr_threshold = 1,
                      pcoa_fill = response_colors,
                      ) 
    res_list[[metric_name]] <- res$plots
    pval <- res$results %>% mutate(feature = metric_name)
    pvals <- rbind(pvals, pval)
  }
```



```{r}
pvals %>% rename(p_value = pvals) %>% 
          mutate(FDR = p.adjust(p_value, "fdr")) %>% 
          select(feature,p_value, FDR)
```

PCoA plotting of samples by response - arrows connect pre-post training samples of the same subject
```{r}
for (metric_name in names(distances_dfs)){
    tmp <- distances_dfs[[metric_name]] %>% left_join(sample_mapping, by = "sample_id")
    bd <- betadisper(d = as.dist(tmp %>% select(tmp$sample_id)), group = unname(tmp$response), bias.adjust = TRUE)
    abd <- anova(bd)
    pbd <- permutest(bd, permutations = 999)
    print(paste("####", metric_name, "####"))
    print(abd)
    print(pbd)
    
    ## Plot distances from group centroid
    ggdata <- data.frame(bd$distances) %>% tibble::rownames_to_column("sample_id") %>% 
                left_join(sample_mapping, by = "sample_id")
      
    ggp <- geom_boxplot_general(ggdata, "response", "bd.distances",color_var = "response",
                           group_var = NULL, palette = response_colors) +
                labs(y = paste(metric_dict_beta[[metric_name]], "\n distance from group center")) +
                custom_theme(is_figure_small = T)
      
    wilcox <- wilcox.test(data = ggdata, bd.distances ~ response, paired = F)
    print(paste(metric_name,"distance from center differences: p = ", wilcox$p.value))      
    
    ggsave(plot = ggp, filename = glue::glue("plots/response_{metric_name}_dist_from_center.png"),
    height = 4, width = 4)
        
    ## Plot PCoA
    arrow_positions <- res_list[[metric_name]][["response_pc1&pc2"]]$data %>% 
                       select(subject, timepoint,response, Axis.1, Axis.2) %>% 
                       tidyr::pivot_wider(names_from = timepoint, values_from = c(Axis.1, Axis.2)) %>% 
                       na.omit()
    
    ggp <- res_list[[metric_name]][["response_pc1&pc2"]] + 
           labs(fill = "", title = "", color = "") + 
           scale_fill_manual(values = response_colors) + 
      geom_curve(data = arrow_positions, 
                aes(x = Axis.1_PRE, xend = Axis.1_POST,
                    y = Axis.2_PRE, yend = Axis.2_POST),
                arrow = arrow(length=unit(0.2,"cm")), size = 0.25) 
    plot(ggp)
    ggsave(plot = ggp + custom_theme() ,height = 4, width = 4,  filename = glue::glue("plots/pcoa_{metric_name}.png"))
    
    

}
legend <- ggp %>% cowplot::get_legend() %>% ggpubr::as_ggplot()
#ggsave(plot = legend ,height = 4, width = 4,  filename = glue::glue("plots/pcoa_legend.png"))
```

```{r}
ggdata <- data.frame(bd$distances) %>% tibble::rownames_to_column("sample_id") %>% 
          left_join(sample_mapping, by = "sample_id") %>% rename(bd_dist = "bd.distances")

geom_boxplot_general(ggdata, "response", "bd_dist",color_var = "response",
                     group_var = NULL, palette = response_colors) +
          labs(y = metric_dict_beta[[metric_name]]) +
          custom_theme(is_figure_small = T)

wilcox.test(data = ggdata, bd_dist ~ response, paired = F)
```


```{r}
df_list <- list()
for (metric_name in names(distances_dfs)){
  df_list[[metric_name]] <-  distances_dfs[[metric_name]] %>% 
         tidyr::pivot_longer(cols = -sample_id, names_to = "sample_id_2", values_to = metric_name) 
}
```

```{r}
beta_div_all <- df_list %>% purrr::reduce(left_join, by = c("sample_id", "sample_id_2"))

samples <- beta_div_all$sample_id %>% unique()
sample_pairs_unique <- combn(samples, 2)%>% t() %>% as.data.frame() %>% 
                       rename(sample_id = V1,  sample_id_2 = V2)

beta_div_all <- beta_div_all %>% inner_join(sample_pairs_unique, by = c("sample_id", "sample_id_2")) %>% 
            mutate(subject_1 = stringr::str_extract(sample_id,"ARE_[0-9]*" ),
                  subject_2 = stringr::str_extract(sample_id_2,"ARE_[0-9]*"),
                  subject_1 = gsub("ARE", "subject", subject_1),
                  subject_2 = gsub("ARE", "subject", subject_2),
                  timepoint_1 = stringr::str_extract(sample_id,"PRE|POST"),
                  timepoint_2 = stringr::str_extract(sample_id_2,"PRE|POST"))

beta_div <- beta_div_all %>% 
            left_join(metadata %>% select(subject, response), by = c("subject_1" = "subject")) %>% 
            left_join(metadata %>% select(subject, response), by = c("subject_2" = "subject")) %>% 
            filter(response.x == response.y) %>% rename(response = response.x)
```

Calculate distance between subject pairs within a response group
```{r}
beta_div_within_tp <- beta_div %>% filter(timepoint_1 == timepoint_2)
res <- data.frame(matrix(nrow = 0, ncol = 4))
colnames(res) <- c("metric", "timepoint", "effsize", "p_value")

for (tp in c("PRE", "POST")){
  ggp_list <- list()
  for (metric_name in names(distances_dfs)){
    ggdata <- beta_div_within_tp %>% filter(timepoint_1 == tp)
    
    # Calculate wilcox manullay for multiple testing correction
    wilcox <- wilcox.test(data = ggdata, as.formula(paste0(metric_name,"~response")), paired = F)
    wilcox_r <- rstatix::wilcox_effsize(data = ggdata, as.formula(paste0(metric_name,"~response")),paired = F )
    res[nrow(res)+1, ] <- c(metric_name, tp,wilcox_r$effsize[[1]], wilcox$p.value)
    
    # Plot pairwise beta diversity between response groups across timepoint and beta diversity metrics
    ggp <- geom_boxplot_general(ggdata, "response", metric_name,
                               group_var = NULL, palette = response_colors) +
          #labs(title = paste(metric_dict_beta[[metric_name]], tolower(tp), "training")) +
          custom_theme(is_figure_small = T)
    ggp_list[[paste(metric_name)]] <- ggp
    ggsave(plot = ggp, filename = glue::glue("plots/response_{metric_name}_{tp}_training.png"),
           height = 4, width = 4)
  }
  margin = theme(plot.margin = unit(c(0.5,0.5,0.5,0.5), "cm"))
  ggp <- gridExtra::grid.arrange(grobs = lapply(ggp_list, "+", margin), ncol = 2)
}
res$p_value_adj <- p.adjust(res$p_value, method = "fdr")
res
```

## Figures 3B, 3C, 3D and 3E - alpha diversity correlation to RMR/TDEE adaptation
```{r}
alpha_diversity <- shannon %>% left_join(metadata_subject_level, by = "subject")
alpha_diversity %>% head()
```

```{r}
alpha_div_delta <- alpha_diversity %>% select(subject, timepoint, shannon) %>% 
          tidyr::pivot_wider(names_from = "timepoint", values_from = c("shannon")) %>% 
          mutate(shannon_delta = post - pre) %>% 
          select(subject, shannon_delta)
```

```{r}
ggdata <- alpha_diversity %>% filter(timepoint == "pre") %>% 
          left_join(alpha_div_delta, by = "subject")
```

```{r, fig.width= 4, fig.height = 4}
features <- c("tdee_adaptation", "rmr_adaptation", "vo2max_delta")
for (alpha_div in c(names(metric_dict_alpha))){
  for (feature in features){
    ggp <- geom_point_with_corr(ggdata, alpha_div, feature, color_var = "response") + 
            labs(x = paste(metric_dict_alpha[[alpha_div]],"index"),
                 y = feature_dict[[feature]],
                 color = "Response\nto training") + 
           scale_color_manual(values = response_colors) +
        custom_theme(is_figure_small = T)
    plot(ggp)
    ggsave(plot = ggp, height = 4, width = 4,
           filename = glue::glue("plots/{alpha_div}_{feature}_corr.png"))
  }
}
```

